---
title: "Florida Coral Reef Report Reproduction"
format: html
---

This example reproduces a figure style from the Florida Coral Reef (FCR) water
quality reporting: **split violin plots** that compare the distribution of each
analyte between the **wet season** and the **dry season**.

* Each violin is split down the middle: the left half is the dry season, the
  right half is the wet season.
* Violins are grouped along the x-axis by **Data Provider** (the SEACAR
  `ProgramName`, mapped to its short name via
  `mapProgramNameToShortName()`).
* There is **one plot (one set of violins) per analyte**.

Seasons follow the South Florida convention:

| Season | Months            |
|--------|-------------------|
| Wet    | May – October     |
| Dry    | November – April  |

```{r setup}
#| code-summary: load libraries and helpers
#| message: false
#| warning: false
library(tidyverse)
library(here)
library(lubridate)

# split violin geom + program-name -> short-name mapping (both shipped in this repo)
source(here("R/GeomSplitViolin.R"))
source(here("R/mapProgramNameToShortName.R"))

# minimum number of observations a (Data Provider x Season) group must have
# before it is drawn as a violin
MIN_OBS <- 10
```

```{r load-data}
#| code-summary: read data, assign season, label Data Provider
df <- readr::read_delim(
  here("data/Discrete WQ - 10006.txt"),
  delim = "|",
  show_col_types = FALSE
) %>%
  mutate(
    DataProvider = mapProgramNameToShortName(ProgramName),
    month_num    = lubridate::month(SampleDate),
    Season       = if_else(month_num >= 5 & month_num <= 10, "Wet", "Dry"),
    Season       = factor(Season, levels = c("Dry", "Wet"))
  ) %>%
  filter(!is.na(ResultValue), is.finite(ResultValue))

cat("Rows:", nrow(df), "\n")
cat("Data Providers:", paste(sort(unique(df$DataProvider)), collapse = ", "), "\n")
cat("Analytes:", dplyr::n_distinct(df$ParameterName), "\n")
```

```{r trim-sparse-groups}
#| code-summary: keep only (analyte, provider) pairs with enough data in both seasons
group_counts <- df %>%
  count(ParameterName, DataProvider, Season, name = "n")

# a provider is kept for an analyte only if BOTH seasons clear MIN_OBS
keep <- group_counts %>%
  pivot_wider(names_from = Season, values_from = n, values_fill = 0) %>%
  filter(Dry >= MIN_OBS, Wet >= MIN_OBS) %>%
  select(ParameterName, DataProvider)

plot_df <- df %>%
  semi_join(keep, by = c("ParameterName", "DataProvider"))

# analytes that still have at least one provider to show
analytes <- plot_df %>%
  distinct(ParameterName) %>%
  arrange(ParameterName) %>%
  pull(ParameterName)

cat("Analytes plotted:", length(analytes), "\n")
```

```{r plot-fn}
#| code-summary: split-violin plot for a single analyte
season_colors <- c(Dry = "#C49A6C", Wet = "#3C8DBC")

# fraction of EACH (Data Provider x Season) distribution hidden off each end of
# the y-axis. The tails are only clipped from *view* (coord_cartesian) -- every
# value still feeds the violin's density estimate, so shapes stay honest.
TRIM_TAILS <- 0.10

plot_analyte <- function(analyte) {
  d <- plot_df %>% filter(ParameterName == analyte)
  unit <- d$ParameterUnits[1]

  # y-window: from the lowest per-group 10th pctile to the highest per-group
  # 90th pctile, so no violin's central mass falls entirely outside the view
  bounds <- d %>%
    group_by(DataProvider, Season) %>%
    summarise(
      lo = quantile(ResultValue, TRIM_TAILS,     na.rm = TRUE),
      hi = quantile(ResultValue, 1 - TRIM_TAILS, na.rm = TRUE),
      .groups = "drop"
    )
  ylim <- c(min(bounds$lo), max(bounds$hi))

  pct <- round(100 * TRIM_TAILS)

  ggplot(d, aes(x = DataProvider, y = ResultValue, fill = Season)) +
    geom_split_violin(
      nudge = 0.03,
      scale = "width",
      trim  = TRUE,
      colour = "grey30",
      alpha = 0.85
    ) +
    stat_summary(
      aes(group = Season),
      fun = median,
      geom = "point",
      position = position_dodge(width = 0.25),
      size = 1.6,
      colour = "grey15",
      show.legend = FALSE
    ) +
    scale_fill_manual(values = season_colors, drop = FALSE) +
    coord_cartesian(ylim = ylim) +
    labs(
      title = analyte,
      subtitle = glue::glue(
        "y-axis clipped to the central {100 - 2 * pct}% range; ",
        "violins computed from all data; median = grey dot"
      ),
      x = "Data Provider",
      y = if (!is.na(unit)) paste0(analyte, " (", unit, ")") else analyte,
      fill = "Season"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "top",
      panel.grid.major.x = element_blank()
    )
}
```

```{r render-all}
#| code-summary: one set of violins per analyte
#| results: asis
#| fig-width: 9
#| fig-height: 5
for (analyte in analytes) {
  cat("\n\n## ", analyte, "\n\n", sep = "")
  print(plot_analyte(analyte))
  cat("\n\n")
}
```

---

**Notes**

* Split-violin drawing comes from `R/GeomSplitViolin.R` (`geom_split_violin()`),
  which is also exported by this package.
* `scale = "width"` makes every violin the same maximum width so shape is
  comparable across providers with very different sample sizes; the grey dot
  marks each season's median.
* `(Data Provider x Season)` groups with fewer than `MIN_OBS`
  (`{r} MIN_OBS`) observations are dropped, and a provider is only shown for an
  analyte when *both* seasons clear that threshold.
* **Tail handling (zoom, not drop).** Analytes here are strongly right-skewed,
  and a few extreme values otherwise stretch the y-axis until the violins
  collapse to lines. Every value still feeds each violin's density estimate, but
  the y-axis is clipped with `coord_cartesian()` to
  `[min group P{r} round(100 * TRIM_TAILS), max group P{r} round(100 * (1 - TRIM_TAILS))]`
  across the `(Data Provider x Season)` groups (`TRIM_TAILS <- {r} TRIM_TAILS`).
  So the shapes stay honest and only the tails are out of frame. Using `ylim()`
  or `scale_y_continuous(limits=)` instead would *discard* those points before
  the density is fit and distort the violins. Set `TRIM_TAILS <- 0` for the full
  range.

**Other ways to handle the long tails** (swap into `plot_analyte()` as needed):

* **Trim instead of zoom** — filter each group to its P10–P90 before plotting;
  the violin then describes only the retained 80% (smaller n, no tail info).
* **Winsorize** rather than trim — `pmin(pmax(x, lo), hi)` piles the clipped mass
  into a spike at each cap, which keeps the sample size intact and signals that a
  tail exists.
* **Compress the axis** — `scale_y_continuous(trans = "sqrt")` (used elsewhere in
  this repo) or `trans = scales::pseudo_log_trans()` tame skew while still
  showing the tails, and both tolerate zeros (`pseudo_log` also tolerates
  negatives).
* **Different mark** — a letter-value / boxen plot (`lvplot::geom_lv()`) or a
  plain boxplot conveys skewed, heavy-tailed data more legibly than a violin.
* **Facet with free y** — `ggforce::facet_row(~ DataProvider, scales = "free_y")`
  gives each provider its own y-range so no single noisy provider flattens the
  rest.
